A Comparison of Parallel Graph Processing Implementations

### 2017-05-04 papers/publication/paper-pollard.tex
OK, final EuroPar 2017 commit.
Updated paper to IEEE format.
It's in a pretty rough transitionary period right now.

\documentclass{llncs}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{hyperref}
\usepackage{cite} % Multiple citations in one \cite command
\usepackage{pgfplotstable}
\usepackage{array} % Required for pgfplotstable dec sep align
\pgfplotsset{compat=1.12}
\usepackage{booktabs}
\usepackage{todonotes}

@@

\begin{document}
\title{A Comparison of Parallel Graph Processing Benchmarks}
\author{Samuel D. Pollard \and Boyana Norris}
\institute{University of Oregon \\
	Eugene OR 97403, USA \\
	\email{spollard@cs.uoregon.edu} \\
	\email{norris@cs.uoregon.edu}
}
\maketitle
\begin{abstract}
The rapidly growing number of large network analysis problems has led to the emergence of many parallel and distributed graph processing systems---one survey in 2014 identified over 80. Since then, the landscape has evolved; some packages have become inactive while more are being developed. Determining the best approach for a given problem is infeasible for most developers. To enable easy, rigorous, and repeatable comparison of the capabilities of such systems, we present an approach and associated software for analyzing the performance and scalability of parallel, open-source graph libraries. We demonstrate our approach on five graph frameworks: GraphMat, the Graph500, the Graph Algorithm Platform, GraphBIG, and PowerGraph using synthetic and real-world datasets. We examine previously overlooked aspects of parallel graph processing performance such as phases of execution and energy usage for three algorithms: breadth first search, single source shortest paths, and PageRank and compare our results to Graphalytics.
\end{abstract}

@@

From a graph algorithm user's point of view, there are optimizations for each system which may not be apparent. For example, the Parallel Boost Graph Library (PBGL)~\cite{Gregor:2005:PBGL} provides generic implementations of their algorithms and the programmer must provide the template specializations. The optimal data structures may differ across algorithms and graphs and must be determined by the programmer.

%% How do we address it?
To mitigate the inherent bias in hand-generated implementations, all the experiments performed use the author-provided implementations; we assume the developers of each system will provide the most performant implementation. While this limits the scope of the experiments it mitigates the bias inherent in our programming skills.

@@

Our research contributions in this paper can be summarized as follows.
\begin{itemize}
	\item Automation of the installation, configuration, and dataset conversion to ensure the experiments execute in the same manner. Essentially, we provide a ``level playing field'' for graph algorithms.
	\item Comparison of algorithm runtime and scalability using several data sets.
	\item Comparison of our framework with the Graphalytics \cite{Capota:2015:Graphalytics} project.
	% BN: I think the bullet below is a bit too low-level for this summary, moving to later section.
	%\item Inspection of the source code of the surveyed parallel graph processing systems to ensure the same phases of execution are measured across differing execution and programming paradigms.
	\item An experimental framework distinguishing between the data structure generation and algorithm execution phases of each system.
	\item Analysis of energy consumption during the algorithm execution phase.
\end{itemize}

\section{Related Work}
A related project is Graphalytics \cite{Capota:2015:Graphalytics}, which also attempts to automate the setup and execution of graph packages for performance analysis. However, in order for Graphalytics to analyze a graph processing system---Graphalytics calls these platforms---a programmer must implement Java classes as wrappers to call the particular implementations, satisfy dependencies, and set the correct shell commands to execute. These all require some knowledge of the inner workings of each system in addition to familiarity with the Graphalytics API and involve a nontrivial amount of programming.

@@

Satish et al.~\cite{Satish:2014:NavigatingGraph} analyze the performance of several systems on datasets on the order of $30$ billion edges and \cite{Lu:2014:ExperimentalEval} uses six real-world datasets and focuses on the vertex-centric programming model.

@@

\section{Experimental Setup}
We refer to a graph's \emph{scale} when describing the size of the graph. Specifically, a graph with scale $S$ has $2^S$ vertices. The datasets are described further in Sect.~\ref{sec:datasets}. We measure strong scaling and speedup by varying the number of threads from one to the total number of threads on our server, 72.

@@

This study explores four shared memory parallel graph processing platforms and one distributed memory framework. The first three are so-called ``reference implementations'' while the remaining two are included because of their performance and popularity. Other popular libraries such as the Parallel Boost Graph Library \cite{Gregor:2005:PBGL} are not considered here because the authors do not provide reference implementations. We consider the following systems.
\begin{itemize}
	\item The Graph500\footnote{We used the most recent version from \url{https://github.com/graph500/graph500}, most similar to release 2.1.4.} \cite{Murphy:2010:Graph500}, a set of large-scale graph benchmarks.
	\item The Graph Algorithm Platform (GAP) Benchmark Suite \cite{Beamer:2015:GAPBench}.
	\item GraphBIG \cite{Nai:2015:Graphbig} benchmark suite.
	\item GraphMat \cite{Sundaram:2015:GraphMat}, a high-performance graph library.
	\item PowerGraph \cite{Gonzalez:2012:Powergraph}, a framework for distributed graph-parallel computation on natural graphs.
\end{itemize}

Our approach is not specific or limited to these graph packages and can be extended to others with relatively modest effort (building each package and if necessary, instrumenting the different phases of the computation).

@@

PageRank is selected because of its popularity; most libraries provide reference implementations. One challenge with using PageRank is the stopping criterion; this is addressed in Sect.~\ref{sec:perf}. Verification of the PageRank results is beyond the scope of this paper, although this may explain some of the large performance discrepancies.

@@

The Graphalytics results in Table~\ref{tab:graphalytics} were performed on the Dota-League dataset. This dataset contains 61,670 vertices and 50,870,313 edges. This dataset comes from the Game Trace Archive\cite{Guo:2012:GTA} and is modified for Graphalytics\footnote{This dataset is available at \url{https://atlarge.ewi.tudelft.nl/graphalytics/}.}. This dataset is useful because it is both weighted and more dense than the usual real-world dataset with an average out-degree of 824.

@@

We analyze the performance of the algorithms described in Sec.~\ref{sec:algs} in terms of execution time, scalability over multiple threads, and power and energy consumption. All figures except those measuring scalability (Fig.~\ref{fig:bfs-scaling}) use 32 threads.

@@

In Table~\ref{tab:graphalytics} we show the results from running Graphalytics on a Kronecker graph of scale $22$. An explanation of each algorithm is given in \cite{Iosup:2016:Graphalyticstech}. Graphalytics by default does not perform SSSP on unweighted, undirected graphs. The discrepancy between PageRank values in Table~\ref{tab:graphalytics} and Fig.~\ref{fig:pr} is a result of the differing stopping criterion.

@@

\begin{figure}
	\centering
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0pt 18pt 18pt 0pt, clip]{graphics/pr_time.pdf}
	\end{minipage}
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0 18pt 18pt 0pt, clip]{graphics/pr_iters.pdf}
	\end{minipage}
	\caption{The $y$-axis is logarithmic only for the left figure. GraphMat continues to run until none of the vertices' ranks change. For the others, we use the stopping condition that the sum of the changes in the weights is no more than $6 \times 10^{-8}$, or approximately machine epsilon for single precision floating point numbers.}
	\label{fig:pr}
\end{figure}

@@

\begin{table}
	\caption{The data are generated using a Kronecker graph with a scale of 22. Sleeping Energy refers to the power (in Watts) consumed during the \texttt{unistd} C \texttt{sleep} function, multiplied by the Time row. Essentially, this measures the amount of energy that would have been consumed if nothing was running. The increase over sleep is the ratio of the first and third columns. These are all averaged over 32 roots.}
	\centering
	\begin{tabular}{l|r|r|r|r}
			&	GAP  &    Graph500 & GraphBIG & GraphMat \\ \hline
		Time (s) &  0.01636 & 0.01884 & 1.600 & 1.424 \\
		Average Power per Root (W) & 72.38 & 97.17 & 78.01 & 70.12 \\
		Energy per Root (J) &	1.184 & 1.830 & 112.213 & 111.104 \\
		Sleeping Energy (J) & 0.4046  & 0.4660 & 39.591 &  35.234 \\
		Increase over Sleep & 2.926 & 3.928 & 2.834 & 3.153
	\end{tabular}
	\label{tab:power}
\end{table}

RAPL also allows the measurement of DRAM power, the results of which are shown in Fig.~\ref{fig:power}. The left plot describes the second row of Table~\ref{tab:power} in more detail. We notice a smaller spread of RAM power consumption, but still a noticeable difference. GraphMat exhibits the lowest average RAM power consumption (but not the shortest execution time). This information can be useful when choosing an algorithm to use in limited power scenarios, where a slower algorithms that will not exceed the power cap is preferred to a faster one that may exceed it.

@@

\section{Future Work and Conclusion}
We have presented a systematic approach to analyzing parallel graph processing performance, both measuring power consumption and the two fundamental phases of execution. While the comparison presented here can help one choose among alternatives for the selected packages and algorithms, the problem of selecting a graph processing framework for a given large-scale problem remains far from simple. Hence, we make our code available at \url{https://github.com/HPCL/easy-parallel-graph} and encourage further experimentation. Our results required minor changes to the original projects to add calls to PAPI sampling and those changes are also available on forked versions of the repositories.

>>>>

\documentclass[conference]{IEEEtran}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{hyperref}
\usepackage{cite} % Multiple citations in one \cite command
\usepackage{pgfplotstable}
\usepackage{array} % Required for pgfplotstable dec sep align
\pgfplotsset{compat=1.12}
\usepackage{booktabs}
\usepackage{todonotes}

@@

\begin{document}
\title{A Comparison of Parallel Graph Processing Implementations}
\author{
	\IEEEauthorblockN{Samuel D. Pollard}
	\IEEEauthorblockA{Computer and Information Science\\
		University of Oregon \\
		Eugene, OR 97403\\
		Email: spollard@cs.uoregon.edu}
	\and
	\IEEEauthorblockN{Boyana Norris}
		\IEEEauthorblockA{Computer and Information Science\\
		University of Oregon \\
		Eugene, OR 97403\\
		Email: norris@cs.uoregon.edu}}
\maketitle
\begin{abstract}
The rapidly growing number of large network analysis problems has led to the emergence of many parallel and distributed graph processing systems---one survey in 2014 identified over 80. Since then, the landscape has evolved; some packages have become inactive while more are being developed. Determining the best approach for a given problem is infeasible for most developers. To enable easy, rigorous, and repeatable comparison of the capabilities of such systems, we present an approach and associated software for analyzing the performance and scalability of parallel, open-source graph libraries. We demonstrate our approach on five graph frameworks: GraphMat, the Graph500, the Graph Algorithm Platform Benchmark Suite, GraphBIG, and PowerGraph using synthetic and real-world\todo{actual real-world. (i) Michael LeBeane et al, Supercomputing (SC 15), Data Partitioning Strategies for Graph Workloads on Heterogeneous Clusters (ii) Shuang Song et al,  ICPP 2016, Proxy-Guided Load Balancing of Graph Processing Workloads on Heterogeneous Clusters.} datasets. We examine previously overlooked aspects of parallel graph processing performance such as phases of execution and energy usage for three algorithms\todo{more algorithms, justify why these algorithms}: breadth first search, single source shortest paths, and PageRank and compare our results to Graphalytics.
\end{abstract}

@@

From a graph algorithm user's point of view, optimizations for each system may not be apparent. For example, the Parallel Boost Graph Library (PBGL)~\cite{Gregor:2005:PBGL} provides generic implementations of their algorithms and the programmer must provide the template specializations. The optimal algorithms may differ across graphs and must be determined by the programmer. To complicate matters further, the $\Delta$-stepping implementation of the parallel Dijkstra's Algorithm for computing single source shortest paths (SSSP) is parametrized by some $\Delta > 0$ for which requires tuning for a given architecture and graph structure to achieve optimal performance \cite{Panitanarak:2014:SSSPPerf}.\todo{Will this be mentioned again?}

We aim to simplify this excess of decisions by both providing results from our experiments on real world and synthetic datasets. We take inspiration from the Graph500 benchmark \cite{Murphy:2010:Graph500} which clearly specifies every step of the breadth first search (BFS) algorithm and how it should be timed. The Graph500 can be used to fairly rank parallel graph processing platforms. We define platforms \todo{as graphalytics does? Or not.} as the combination of hardware, operating system, and graph processing library. We refer the reader to Section~[SECTION] for the platforms used in this paper. Thus, a potential user of parallel graph algorithms should have no difficulty selecting the optimal library given a wide variety of graphs and performance data available on his or her system.

We select a small number of graph processing platforms over which we attempt to homogenize all aspects of execution to truly compare performance.

%% How do we address it?
All the experiments performed use the author-provided implementations with modifications only to insert performance analysis hooks or to ensure homogeneous termination criteria; we assume the developers of each system will provide the best performing implementation. While this limits the scope of the experiments it mitigates the bias inherent in our programming skills.\todo{Should I talk about how library designers are biased?}

@@

Our research contributions in this paper can be summarized as follows.
\begin{itemize}
	\item Automation of the installation\todo{This is sort of done in the repository, but not really explained in the paper}, configuration, and dataset conversion to ensure the experiments execute in the same manner. Essentially, we provide a ``level playing field'' for graph algorithms.\todo{put URL to github, make public}
	\item Comparison of algorithm runtime and scalability using several data sets.
	\item Comparison of our framework with the Graphalytics \cite{Capota:2015:Graphalytics} project.
	% BN: I think the bullet below is a bit too low-level for this summary, moving to later section.
	%\item Inspection of the source code of the surveyed parallel graph processing systems to ensure the same phases of execution are measured across differing execution and programming paradigms.
	\item An experimental framework distinguishing between the data structure generation and algorithm execution phases of each system.
	\item Analysis of energy consumption during the algorithm execution phase.
\end{itemize}
\todo{Clarify that we're not introducing our own benchmark suite instead we're comparing existing frameworks at a greater depth.}

\section{Related Work}
\todo{Introduction sentence. How many others are there? There's a large number of qualatative. Most third party comparisons are qualitative and just describe the overview instead of actual runtimes. This isn't a level playing field.}
A related project is Graphalytics \cite{Capota:2015:Graphalytics}, which also attempts to automate the setup and execution of graph packages for performance analysis. However, in order for Graphalytics to analyze a graph processing system---Graphalytics calls these platforms---a programmer must implement Java classes as wrappers to call the particular implementations, satisfy dependencies, and set the correct shell commands to execute. These all require some knowledge of the inner workings of each system in addition to familiarity with the Graphalytics API and involve a nontrivial amount of programming.

\todo{Cite the Exploring big graph computing - An empirical study from architectural perspective, but we dont do an architectural perspective. Don't care about distinctino between framework/usercode since we're just using the vendor provided implementations.}
Satish et al.~\cite{Satish:2014:NavigatingGraph} analyze the performance of several systems on datasets on the order of $30$ billion edges and \cite{Lu:2014:ExperimentalEval} uses six real-world datasets and focuses on the vertex-centric programming model.

@@

@@

\section{Methodology}

\todo{Start with dataset subsection, write new intro to methodology, move this into inputs. We're comparing algorithms not applications. Understand those since they're the building blocks. Not comparing architectural details (cite that Nai paper). Important paragraph.}
We refer to a graph's \emph{scale} when describing the size of the graph. Specifically, a graph with scale $S$ has $2^S$ vertices. \todo{Say 2 to the S = 2351501351035923} The datasets are described further in Sect.~\ref{sec:datasets}. We measure strong scaling and speedup by varying the number of threads from one to the total number of threads on our server, 72.

@@

This study explores four shared memory parallel graph processing platforms and one distributed memory framework. The first three are so-called ``reference implementations'' while the remaining two are included because of their performance and popularity. Other popular libraries such as the Parallel Boost Graph Library \cite{Gregor:2005:PBGL} are not considered here because the authors do not provide reference implementations. We consider the following systems.
\begin{itemize}
	\item The Graph500\footnote{We used the most recent version from \url{https://github.com/graph500/graph500}, most similar to release 2.1.4.} \cite{Murphy:2010:Graph500}, a set of large-scale graph benchmarks.
	\item The Graph Algorithm Platform (GAP) Benchmark Suite \cite{Beamer:2015:GAPBench}.
	\item GraphBIG \cite{Nai:2015:Graphbig} benchmark suite.
	\item GraphMat \cite{Sundaram:2015:GraphMat}, a high-performance graph library.
	\item PowerGraph \cite{Gonzalez:2012:Powergraph}, a framework for distributed graph-parallel computation on natural graphs.
	\todo{Explain each more.}
\end{itemize}

Our approach is not specific or limited to these graph packages and can be extended to others with relatively modest effort (building each package and if necessary, instrumenting the different phases of the computation).
\todo{Say what we automate and what we don't automate.}
\todo{Weighted versus unweighted}

@@

PageRank is selected because of its popularity; most libraries provide reference implementations. One challenge with using PageRank is the stopping criterion; this is addressed in Sect.~\ref{sec:perf}. Verification of the PageRank results is beyond the scope of this paper, although this may explain some of the large performance discrepancies.\todo{Better explain why these were selected. They're the most popular of the algorithms for the papers we surveyed.}

@@

The Graphalytics results in Table~\ref{tab:graphalytics} were performed on the Dota-League dataset\todo{One reviewer confused this dataset with the Kronecker.}. This dataset contains 61,670 vertices and 50,870,313 edges. This dataset comes from the Game Trace Archive\cite{Guo:2012:GTA} and is modified for Graphalytics\footnote{This dataset is available at \url{https://atlarge.ewi.tudelft.nl/graphalytics/}.}. This dataset is useful because it is both weighted and more dense than the usual real-world dataset with an average out-degree of 824.

@@

We analyze the performance of the algorithms described in Sec.~\ref{sec:algs} in terms of execution time, scalability over multiple threads, and power and energy consumption\todo{ A precise list and exposition of the requirements and changes	done to the application would help in understanding the methodology and its implications}. All figures except those measuring scalability (Fig.~\ref{fig:bfs-scaling}) use 32 threads.

@@

In Table~\ref{tab:graphalytics} we show the results from running Graphalytics on a Kronecker graph of scale $22$ (4,194,304 vertices and approximately 33,554,432 edges)\todo{work?}. An explanation of each algorithm is given in \cite{Iosup:2016:Graphalyticstech}. Graphalytics by default does not perform SSSP on unweighted, undirected graphs. The discrepancy between PageRank values in Table~\ref{tab:graphalytics} and Fig.~\ref{fig:pr} is a result of the differing stopping criterion.\todo{What are the likely culprits for performance differences---measure cache miss rates. Try to find something which correlates with performance. Resource stalls will work easily (not as interesting) Cycles per instruction? Automate with perfexplorer or autoperf}

@@

\todo{The y-axis is mislabeled. Should be "iterations" for pagerank.}
\begin{figure}
	\centering
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0pt 18pt 18pt 0pt, clip]{graphics/pr_time.pdf}
	\end{minipage}
	\begin{minipage}{0.48\linewidth}
		\includegraphics[width=\linewidth, trim=0 18pt 18pt 0pt, clip]{graphics/pr_iters.pdf}
	\end{minipage}
	\caption{The $y$-axis is logarithmic only for the left figure. GraphMat continues to run until none of the vertices' ranks change. For the others, we use the stopping condition that the sum of the changes in the weights is no more than $6 \times 10^{-8}$, or approximately machine epsilon for single precision floating point numbers.}
	\label{fig:pr}
\end{figure}

@@

\begin{table}
	\caption{The data are generated using a Kronecker graph with a scale of 22 and using 32 threads. Sleeping Energy refers to the power (in Watts) consumed during the \texttt{unistd} C \texttt{sleep} function, multiplied by the Time row. Essentially, this measures the amount of energy that would have been consumed if nothing was running. The increase over sleep is the ratio of the first and third columns. These are all averaged over 32 roots.}
	\centering
	\begin{tabular}{l|r|r|r|r}
			&	GAP  &    Graph500 & GraphBIG & GraphMat \\ \hline
		Time (s) &  0.01636 & 0.01884 & 1.600 & 1.424 \\
		Average Power per Root (W) & 72.38 & 97.17 & 78.01 & 70.12 \\
		Energy per Root (J) &	1.184 & 1.830 & 112.213 & 111.104 \\
		Sleeping Energy (J) & 0.4046  & 0.4660 & 39.591 &  35.234 \\
		Increase over Sleep & 2.926 & 3.928 & 2.834 & 3.153
	\end{tabular}
	\label{tab:power}
\end{table}

RAPL also allows the measurement of DRAM power, the results of which are shown in Fig.~\ref{fig:power}. The right plot describes the second row of Table~\ref{tab:power} in more detail. We notice a smaller spread of RAM power consumption, but still a noticeable difference. GraphMat exhibits the lowest average RAM power consumption (but not the shortest execution time). This information can be useful when choosing an algorithm to use in limited power scenarios, where a slower algorithms that will not exceed the power cap is preferred to a faster one that may exceed it.

@@

\section{Future Work and Conclusion}
\todo{Maybe; it's still difficult to add more graphs to the dataset sinc you need to know all the metadata.}
We have presented a systematic approach to analyzing parallel graph processing performance, both measuring power consumption and the two fundamental phases of execution. While the comparison presented here can help one choose among alternatives for the selected packages and algorithms, the problem of selecting a graph processing framework for a given large-scale problem remains far from simple. Hence, we make our code available at \url{https://github.com/HPCL/easy-parallel-graph} and encourage further experimentation. Our results required minor changes to the original projects to add calls to PAPI sampling and those changes are also available on forked versions of the repositories.
